---
title: "Untitled"
author: "Miguel"
date: "18/01/2022"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

# Divvy_Exercise_Full_Year_Analysis

This analysis is based on the Divvy case study "'Sophisticated, Clear,
and Polished': Divvy and Data Visualization" written by Kevin Hartman
(found here: <https://artscience.blog/home/divvy-dataviz-case-study>).
The purpose of this script is to consolidate downloaded Divvy data into
a single dataframe and then conduct simple analysis to help answer the
key question:



 Install required packages

 tidyverse for data import and wrangling

 lubridate for date functions

 ggplot for visualization


```{r}
    library(tidyverse)  #helps wrangle data
    library(lubridate)  #helps wrangle date attributes
    library(ggplot2)  #helps visualize data
    getwd() #displays your working directory
    setwd("/Users/maap_/Documents/trip data/TRIP_2021") #sets your working directory to simplify calls to data ... make sure to use your OWN username instead of mine ;)
```
#COLLECT DATA Trips 2021
 Upload Divvy datasets (csv files) here
```{r}
    trip_01 <- read_csv("202101-divvy-tripdata.csv")
    trip_02 <- read_csv("202102-divvy-tripdata.csv")
    trip_03 <- read_csv("202103-divvy-tripdata.csv")
    trip_04 <- read_csv("202104-divvy-tripdata.csv")
    trip_05 <- read_csv("202105-divvy-tripdata.csv")
    trip_06 <- read_csv("202106-divvy-tripdata.csv")
    trip_07 <- read_csv("202107-divvy-tripdata.csv")
    trip_08 <- read_csv("202108-divvy-tripdata.csv")
    trip_09 <- read_csv("202109-divvy-tripdata.csv")
    trip_10 <- read_csv("202101-divvy-tripdata.csv")
    trip_11 <- read_csv("202101-divvy-tripdata.csv")
    trip_12 <- read_csv("202101-divvy-tripdata.csv")
```

# WRANGLE DATA AND COMBINE INTO A SINGLE FILE

#Compare

column names each of the files While the names don't have to be in
the same order, they DO need to match perfectly before we can use a
command to join them into one file.
```{r}
    colnames(trip_01)
    colnames(trip_02)
    colnames(trip_03)
    colnames(trip_04)
```

# Inspect the dataframes and look for incongruencies

```{r}
    str(trip_01)
    str(trip_02)
    str(trip_03)
    str(trip_04)
    str(trip_05)
    str(trip_06)
    str(trip_07)
    str(trip_08)
    str(trip_09)
    str(trip_11)
    str(trip_12)
```

# Stack individual months data frames into one annual big data frame

```{r}
    Trips_2021 <- bind_rows(trip_01,trip_02,trip_03,trip_04,trip_05,trip_06,trip_07,trip_08,trip_09,trip_10,trip_11,trip_12)
```

# Remove lat, long.

```{r}
    Trips_2021 <- Trips_2021 %>% select(-c(start_lat, start_lng, end_lat, end_lng))
```

# CLEAN UP AND
ADD DATA TO PREPARE FOR ANALYSIS

```{r}
    colnames(Trips_2021)  #List of column names
    nrow(Trips_2021)  #How many rows are in data frame?
    dim(Trips_2021)  #Dimensions of the data frame?
    head(Trips_2021)  #See the first 6 rows of data frame.  Also tail(Trips_2021)
    str(Trips_2021)  #See list of columns and data types (numeric, character, etc)
    summary(Trips_2021)  #Statistical summary of data. Mainly for numerics
```

# There are a few problems we will need to fix:

We will want to add some additional columns of data -- such as day, month, year -- that provide additional opportunities to aggregate the data.

```{r}
    Trips_2021$date <- as.Date(Trips_2021$started_at) #The default format is yyyy-mm-dd
    Trips_2021$month <- format(as.Date(Trips_2021$date), "%m")
    Trips_2021$day <- format(as.Date(Trips_2021$date), "%d")
    Trips_2021$year <- format(as.Date(Trips_2021$date), "%Y")
    Trips_2021$day_of_week <- format(as.Date(Trips_2021$date), "%A")
```
# Add a "ride_length" calculation to all_trips (in seconds)
### <https://stat.ethz.ch/R-manual/R-devel/library/base/html/difftime.html>

```{r}
    Trips_2021$ride_length <- difftime(Trips_2021$ended_at,Trips_2021$started_at)
```

# Inspect the structure of the columns
```{r}
    str(Trips_2021)
```

# Convert "ride_length" from Factor to numeric so we can run calculations on the data
```{r}
    is.factor(Trips_2021$ride_length)
    Trips_2021$ride_length <- as.numeric(as.character(Trips_2021$ride_length))
    is.numeric(Trips_2021$ride_length)
```
# Remove "bad" data
 The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative
We will create a new version of the dataframe (v2) since data is being removed
### <https://www.datasciencemadesimple.com/delete-or-drop-rows-in-r-with-conditions-2/>

```{r}
    trips_v2 <- na.omit(Trips_2021[!(Trips_2021$ride_length < 0),])
```
# DESCRIPTIVE ANALYSIS
Ride_length (all figures in seconds)

```{r}

    mean(trips_v2$ride_length) #straight average (total ride length / rides)
    median(trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
    max(trips_v2$ride_length) #longest ride
    min(trips_v2$ride_length) #shortest ride
```
condense the four lines above to one line using summary() on the specific attribute

```{r}
    summary(trips_v2$ride_length)
```
Compare members and casual users

```{r}
    aggregate(trips_v2$ride_length ~ trips_v2$member_casual, FUN = mean)
    aggregate(trips_v2$ride_length ~ trips_v2$member_casual, FUN = median)
    aggregate(trips_v2$ride_length ~ trips_v2$member_casual, FUN = max)
    aggregate(trips_v2$ride_length ~ trips_v2$member_casual, FUN = min)
```

average ride time by each day for members vs casual users
```{r}
    aggregate(trips_v2$ride_length ~ trips_v2$member_casual + trips_v2$day_of_week, FUN = mean)
```
In order
```{r}
    trips_v2$day_of_week <- ordered(trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

    aggregate(trips_v2$ride_length ~ trips_v2$member_casual + trips_v2$day_of_week, FUN = mean)
```

# Analyze ridership data by type and weekday

```{r}
    trips_v2 %>% 
      mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
      group_by(member_casual, weekday) %>%  #groups by usertype and weekday
      summarise(number_of_rides = n()                           #calculates the number of rides and average duration 
                ,average_duration = mean(ride_length)) %>%      # calculates the average duration
      arrange(member_casual, weekday)                               # sorts
```

# Let's visualize the number of rides by rider type

```{r}
    trips_v2 %>% 
      mutate(weekday = wday(started_at, label = TRUE)) %>% 
      group_by(member_casual, weekday) %>% 
      summarise(number_of_rides = n() ,average_duration = mean(ride_length)) %>% 
      arrange(member_casual, weekday)  %>% 
      ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
      geom_col(position = "dodge")
```

# Let's create a visualization for average duration

```{r}
    trips_v2 %>% 
      mutate(weekday = wday(started_at, label = TRUE)) %>% 
      group_by(member_casual, weekday) %>% 
      summarise(number_of_rides = n()
                ,average_duration = mean(ride_length)) %>% 
      arrange(member_casual, weekday)  %>% 
      ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
      geom_col(position = "dodge")
```

# EXPORT SUMMARY FILE
# <https://datatofish.com/export-dataframe-to-csv-in-r/>


write.csv(trips_v2,"C:\\Users\\maap\_\\Documents\\trip
data\\TRIP_2021\\Year_trips_2021.csv", row.names = TRUE )
